Predictive Models to Support Quoting of Fixed Fee Consulting Projects

### 2016-12-16 finalwriteup/Consulting Profitability Paper/Consulting_Profitability_Paper.tex
finished method in paper

Neural Netowrks and SVM were not purusued because black box A previously
mentioned disadvantage is that the NN algorithm is a black box because
the internal structure is too complex for interpretation. They also
require a lot of training data relative to other methods.

Also, SVM's can be very slow to train and therefore not suitable for
industry purposes (Auria and Moro 2008). Because it is so important to
engage decision makers with a model that can explain its results, these
two methods are not appropriate for the effort estimation problem.

@@

Because of the linear relationship, an equation can be fit to the log
odds of the binary response variable against each of the explanatory
variables using linear regression. Then, the linear equation describing
the log odds can be transformed back to probability by taking the
inverse log i.e.~the exponential and rearranging (Macdonald 1975):

The result in a sigmoidal function bound by 0 and 1. The sigmoidal
function originates from a linear fit of the log odds in the data.
Assume linear relationship between covariates and response variable

A coefficient represents the change in the log odds of the response
variable for each unit increase in an explanatory variable. Therefore,
taking the exponential of the coefficient is the change in odds of the
response variable. If the change in odds was 2, then if the explanatory
variable increased by 1, the response variable event would be twice as
likely to occur.

Logistic Regression is a good benchmark to compare other binary
predictive models due to its simplicity and speed (Moore and McCabe
1989). single linear relationship means Not prone to overfitting

@@

single decision trees also suffer from low predictive accuracy and
instability, so to combat these problems, ensemble tree methods were
pioneered in the 1990's (Breiman 1996).

The three methods of combining hundreds or thousands of trees turn
decision trees into high performing, stable predictors.

training multiple trees from the same data by sampling bootstrapped
training sets with replacement. However, when creating each tree, a
random subset of attributes (variables) is considered at each split. The
reduced subset of attributes is resampled for each split in the tree.
This allows dominant variables to be suppressed for a fraction of the
splits, allowing the algorithm to explore signals in weaker variables
(Breiman 2001).

Additionally, ensemble tree methods are faster than SVM's and Neural
Networks but can perform just as well and even provide insights such as
variable importance and variable relationships (Sealfon and Gymrek
2012).

@@

Lastly, the boosted decision tree approach applies gradient descent
theory to a series of decision trees. The trees are limited to a certain
depth to maintain simplicity, and each tree models the residuals (or
errors) of the preceding tree. By modeling the errors, misclassified
cases are weighted higher than correctly classified cases and influence
the structure of the latest tree more (Hastie, Tibshirani, and Friedman
2009). This increased weighting on errors is why the algorithm is called
boosting. The limited depth of each tree prevents overfitting at each
stage and the combined result of up to thousands of trees is very
powerful (Elith, Leathwick, and Hastie 2008).

Caruana and Niculescu-Mizil (2006) tested Boosted Trees, Random Forests,
Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary
classification problems and found that Boosted Trees performed best,
followed by Random Forests. This demonstrates how ensemble tree methods
are capable of competing with high-level machine learning algorithms.

Trees are not built on a probabilistic framework, and therefore their
results cannot be provided in this framework (Louppe and Prettenhofer
2014).

@@

The Naive Bayes method works by making conditional independence
assumptions about the explanatory variables in order to simplify
probability calculations for the response variable (the response
variable must be categorical).

These equations are simple to compute given the data, and the class with
the highest probability can then be chosen (F. Provost and Fawcett
2013).

The advantages of this method are that the conditional independence
assumption enables very fast calculations and predictions. The method
can perform well in real world tasks because the assumption of
independence does not significantly damage predictions (F. Provost and
Fawcett 2013). Assumptions of independence genearlly not true
(correlated variables). This is fine for ranking but the output
probabilities are not accurate statistical probabilities (Caruana and
Niculescu-Mizil 2006).

Naive Bayes method generally provides a good benchmark to compare
against more complex models that should outperform it (Caruana and
Niculescu-Mizil 2006).

\subparagraph{3.1.5 Bayesian Networks}\label{bayesian-networks}

A Bayesian Network is a graphical probabilistic model that illustrates
the conditional dependencies between variables in a data set. The model
is visually represented by a directed acyclic graph (DAG) and is capable
of linking the conditional dependency between any variable to another
variable (Heckerman 1998). The conditional relationships are Bayesian,
where the probabilities in one node are conditional upon values in nodes
directed towards it as well as preceding nodes

Bayesian Networks reduce computations required to find the probability
of a unique combination given other explanatory variable values while
still taking into account many conditional dependencies that Naive Bayes
(Barber 2012)

Naive Bayes is the simplest form of a Bayesian Network (Zhang 2004)

Network relationships can be learned from the data, however this is not
widely included as part of the suite of machine learning methods. The
conditional variable dependencies are calculated from the data, which in
turn can define the graph structure. Some conditional dependencies can
be set before the structure is learned (Barber 2012).

Drawbacks of Bayesian Networks include the difficulty for them to
process continuous variables.

Bayesian networks have found success in combining deterministic models
with observed data as well as expert knowledge.(Kragt 2009)

@@

It includes first how the data was obtained, the lengthy cleaning
process, Projects varied from total invoiced amounts of \$500 to over
\$1,000,000 a project could have thousands of rows of relevant data that
needed to be converted into a single row per project. 10 engineered
variables describing timesheets: Percent of hours performed by
`professional' employees as opposed to `technical' employees, Position
of the employee that completed the most hours on each project 3
engineered from invoicing data: 4 from project data: number of projects
completed with each client text analysis of project titles to develop
project classifications

Discretisation was performed by generating a hierarchical dendrogram of
each variable to visualise the clusters.

Limiting the predictive model to a concise set of meaningful variables
reduces noise and improves predictions. Less variables and a simpler
model is easier for stakeholders to understand (Weisberg 2005). followed
by variable importance analysis, variable selection: varialbe importance
aalyses done with linear regression, random forest, cforest to represent
the range of algroithms that were trialled. 11 variables chosen:

trials of selected predictive algorithms. binary classification problem
predicting profit or loss. Complex models should be measured against
simple models that can be built at a fraction of the computational cost.

\begin{itemize}
\tightlist
\item
  Regression - baseline model
\item
  Naive Bayes - baseline model
\item
  Bayesian Network - grabage.
\item
  Random Forest
\item
  Gradient Boosted Trees
\end{itemize}

To compare the models, the area under the receiver operating
characteristic (ROC) curve (AUC) statistic was used. An ROC graph
visualises the curve from which an AUC score is calculated. In this
problem a `positive' is a loss making job. For binary classification,
AUC is a more meaningful statistic than classification accuracy when the
output is a probability that can be applied to the problem. The AUC
indicates model performance across many thresholds while classification
accuracy represents a single threshold.

a model must be first created using a training set. The model can then
make predictions on the test set. this is used to build the ROC A model
that is perfectly classified would have an AUC = 1. A model that
predicts as well as random chance would have an AUC = 0.5. An AUC
between 0.5 and 1 means the model is performing better than random
chance.

In order to compare which models performed significantly better than
others, an adequate sample size of results statistics is required.
Multiple models could be made by using different data in the training
vs.~testing sets, each providing a resulting test statistic (RMSE or
AUC). Initially, 20 models of each method were created in this fashion.
Then a two-sample power calculation was run using the two sets of 20
results to determine the sample size to achieve a statistical power of
0.8 (100 to achieve power of 0.8 between boosted trees and naive bayes)

Missing Data Imputation. All methods except gradient Boosted Trees and
Naive Bayes could not handle missing data. a complete data set allows
for complete sets of predictions from each method, and these predictions
could then be blended to further improve results. The MICE Random Forest
method was chosen for imputation because it has been proven to work well
with complex data sets (Shah et al. 2014). If similar predictive results
were obtained using Boosted Trees imputed data and imputed data, the
imputed data must be reasonable. The imputed data set was then trialed
on the remaining methods and compared to unimputed trials.

Once the best methods were selected, they were blended in numerous ways,
using both simple averaging techniques and sophisticated
machine-learning algorithms. These were compared against individual
models and the best constructs were selected. Six blending methods,
ranging from simple to complex, were tested using predictions from the
top performing individual models. These included simple averaging of the
individual model results, building a Logistic Regression model using the
individual model results only, a Boosted Tree model using individual
model results only, feature weighted linear stacking (FWLS), Random
Forests, and Boosted Trees. A simplified explanation of FWLS is as a
Linear Regression where meta-features as well as model predictions from
individual models are included as explanatory variables. Then, each
meta-feature is interacted with each set of model results (Sill et al.
2009). Feature interaction is performed passively due to the nature of
how trees are built. A split in a node determined by one variable is
conditional upon the preceding split, which was based on another
variable and so on. The simplest method averaged the probability output
(a number between 0 and 1) of all three models. The next two simplest
methods consisted of building a Logistic Regression and Boosted Trees
model from the three prediction model outputs only. Feature Weighted
Linear Stacking (FWLS), Random Forest and Boosted Trees were also tested
as blending methods in a more complex scenario where the predictions
from each model became additional variables to the original explanatory
variables (called meta-features in this context). The model predictions
were interacted with each original variable so that models that
performed more strongly under certain meta-feature states could be
weighted as such.

Again, 100 models were built to achieve a power of 0.8 for the variation
in AUC test statistics across different divisions of training/testing
data. A maximum of five blended models were created from a complete set
of test results (built from the individual models), then a new set of
test results were created for the next five blended models.

Finally the impact of the algorithm on the overall profits of the case
study company was analysed via decision-making scenarios. This analysis
differed from accuracy in predicting profitability, so several blended
models as well as individual models were carried forward for this
analysis. This was done using a profit curve - a chart that plots the
change in profit the company earns on the y-axis vs.~the probability
threshold on the x-axis. A simple approach was taken for this analysis,
where projects with a probability to be a loss making job greater than
the threshold were rejected entirely. Therefore all profits and losses
from jobs above the threshold were discounted. An equation defining the
change in profit as a percentage of the original profit using threshold
rejection is shown below:

\begin{equation}
\Delta\ Overall\ Profit\ \textrm{(\%)} = \frac{\sum_{p = 1}^{N} I(Pr_p < threshold) \cdot Profit_p}{\sum_{p = 1}^{N} Profit_p} \cdot100
\end{equation}

@@

If the threshold was zero, all jobs were rejected and the profit would
be \$0. If the threshold was 1.0, all jobs were accepted and the profit
would be the same as the profit the company actually experienced since
the data is a sample of historic projects. The aim was to find the
optimal threshold point where saying `no' to a job above that level
would result in higher profits, because jobs that were likely to make a
loss were being rejected. This chart will clarify what percentage of
profit increase the company could expect by integrating the algorithm
into decision-making.

the curve varies with different divisions of the data. Therefore, in
order to understand the uncertainty around the profit curve and to
determine which blended or individual model performed statistically
better than others, a large sample size of curves was required. 100 to
achieve a power of 0.8. which was achieved by repeating 5-fold
training/testing splits. A 95\% confidence interval could also be
determined around the highest point on each curve. The final expected
increase in profit and the percentage of projects to be rejected
presented a clear scenario that the case study business managers could
assess in terms of their business strategy.

@@

\hypertarget{ref-Auria2008}{}
Auria, Laura, and Rouslan A Moro. 2008. ``Support Vector Machines (Svm)
as a Technique for Solvency Analysis.'' Journal Article.

\hypertarget{ref-Barber2012}{}
Barber, David. 2012. \emph{Bayesian Reasoning and Machine Learning}.
Book. Cambridge University Press.

@@

\hypertarget{ref-Heckerman1998}{}
Heckerman, David. 1998. \emph{A Tutorial on Learning with Bayesian
Networks}. Book. Springer.

@@

\hypertarget{ref-Kragt2009}{}
Kragt, Marit E. 2009. \emph{A Beginners Guide to Bayesian Network
Modelling for Integrated Catchment Management}. Book. Landscape Logic.

@@

\hypertarget{ref-Louppe2014}{}
Louppe, Gilles, and Peter Prettenhofer. 2014. ``Gradient Boosted
Regression Trees.'' Online Multimedia. PyData.

@@

@@

\hypertarget{ref-Shah2014}{}
Shah, Anoop D, Jonathan W Bartlett, James Carpenter, Owen Nicholas, and
Harry Hemingway. 2014. ``Comparison of Random Forest and Parametric
Imputation Models for Imputing Missing Data Using Mice: A Caliber
Study.'' Journal Article. \emph{American Journal of Epidemiology},
kwt312.

@@

\hypertarget{ref-Zhang2004}{}
Zhang, Harry. 2004. ``The Optimality of Naive Bayes.'' Journal Article.
\emph{AA} 1 (2): 3.

>>>>

A range of simple and complex, linear and non-linear methods were
applied to the problem including Logistic Regression, Random Forests,
Gradient Boosted Trees, and Naiive Bayes. Furthermore, logistic
regression, random forests, boosted trees and simple averaging were used
to blend predictions from individual models. A brief description of each
method is provided below.

@@

Logisitic regression models a linear relationship bewteen the log odds
of a binary response variable and each of the explanatory variables.
This is performed similarly to linear regression, but is transformed to
probability of an event occuring by taking the inverse log of the log
odds and rearranging (Macdonald 1975):

\[
\begin{aligned}
\log(odds) &= \beta_0 + X_1\beta_1 \\
\log\left(\frac{p}{1-p}\right) &= \beta_0 + X_1\beta_1 \\
-\log\left(\frac{1-p}{p}\right) &= \beta_0 + X_1\beta_1 \\
-\log\left(\frac{1}{p} - 1\right) &= \beta_0 + X_1\beta_1 \\
\log\left(\frac{1}{p} - 1\right) &= -\beta_0 - X_1\beta_1 \\
\exp\left(\log\left(\frac{1}{p} - 1\right)\right) &= \exp(-\beta_0 - X_1\beta_1) \\
\frac{1}{p} - 1 &= \exp(-\beta_0 - X_1\beta_1) \\
\frac{1}{p} &= 1 + \exp(-\beta_0 - X_1\beta_1) \\
p &= \frac{1}{1 + \exp(-\beta_0 - X_1\beta_1)}
\end{aligned}
\]

The result in a sigmoidal function bound by 0 and 1. A coefficient
represents the change in the log odds of the response variable for each
unit increase in an explanatory variable. Logistic Regression is a good
benchmark to compare other binary predictive models due to its
simplicity and speed (Moore and McCabe 1989). Addittionally, because of
the simple, single linear relationship, the method is not prone to
overfitting (Perlich, Provost, and Simonoff 2003).

@@

Ensemble tree methods were pioneered in the 1990's to combat the low
predictive accuracy and instability of individual decision trees.
Combining hundreds or thousands of trees transform decision trees into
high performing, stable predictors (Breiman 1996).

The Random Forests algorithm works by training multiple trees from the
bootstrapped samples of a dataset. Additionally, when creating each
tree, a random subset of attributes (covariates) is considered at each
split. The reduced subset of covariates is resampled for each split in
the tree. Dominant covariates are then suppressed for a fraction of the
splits, allowing the algorithm to explore signals in weaker covariates
(Breiman 2001). Performance statistics and tests run within the
algorithm provide insights to the modeller, such as variable importance
and variable relationships (Sealfon and Gymrek 2012).

@@

The boosted decision tree approach applies gradient descent theory to a
series of decision trees. Performance tends to be higher if individual
trees are limited to a certain depth to maintain simplicity and prevent
overfitting (Elith, Leathwick, and Hastie 2008). Unlike Random Forests,
the trees are sequential, where each tree models the residuals (or
errors) of the preceding tree.

By modeling the errors, misclassified cases are weighted higher than
correctly classified cases and influence the structure of the next tree
more (Hastie, Tibshirani, and Friedman 2009). This increased weighting
of error cases is called boosting. The successive results of up to
thousands of trees is very powerful (Elith, Leathwick, and Hastie 2008).
Caruana and Niculescu-Mizil (2006) tested Boosted Trees, Random Forests,
Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary
classification problems and found that Boosted Trees performed best,
followed by Random Forests. This demonstrates how capable ensemble tree
methods are of competing with high-level machine learning algorithms.

@@

Naive Bayes is a simple, fast method included in this work as it
provides a good baseline method against which to compare complex methods
(Caruana and Niculescu-Mizil 2006). It works by making conditional
independence assumptions about the explanatory variables in order to
simplify probability calculations for a categorical response variable.
For example, if predicting a binary variable, \(R\) (with classes 0 or
1) given two covariates \(X_1\) and \(X_2\), the equations for the
probability of both outcomes is as follows:

\(p(R = 1|X_1 X_2) = p(X_1|R = 1) \cdot p(X_2|R = 1) \cdot p(R = 1)\)

\(p(R = 0|X_1 X_2) = p(X_1|R = 0) \cdot p(X_2|R = 0) \cdot p(R = 0)\)

These equations are simple to compute given the data, and the class with
the highest probability can then be chosen (F. Provost and Fawcett
2013). The method can perform well in real world tasks because the
assumption of independence does not significantly damage predictions,
but amplifies the magnitude of the probability (F. Provost and Fawcett
2013). This is fine for ranking class 0 against 1 but the output
probabilities are not accurate statistical probabilities (Caruana and
Niculescu-Mizil 2006).

\subparagraph{3.1.5 Model Blending}\label{model-blending}

Several research groups involved in the high profile data science
competition, Netflix Prize, developed sophisticated methods of model
stacking, otherwise known as ensemble methods or model blending (Sill et
al. 2009). In this paper, the term `blending' will be used to avoid
confusion with ensemble tree methods. Model blending combines
predictions from several models that used methods with different
theoretical foundations. In this way, the strengths of each method can
be combined. Since the 1990's research in statistics has shown that
averaging results from different methods can provide better predictive
accuracy than any single model (Madigan and Raftery 1994).

Six blending methods, ranging from simple to complex, were chosen. These
included simple averaging of the individual model results, building a
Logistic Regression model and Boosted Tree model using individual model
results only, feature weighted linear stacking (FWLS), Random Forests,
and Boosted Trees. The first two models are simply the average, and
weighted average of predictions from separate individual models. For
example, if the Logistic Regression model output a probability of 0.6
for one case, while a Boosted Tree output 0.7, a simple blended average
for that case would be a probability of 0.65. The last three methods are
more complex, and take advantage of potentially meaningful interactions
between the individual model predictions and the original covariates
describing each case (called meta-features in the model blending
literature) (Sill et al. 2009). For example, if the Logistic Regression
model predicted profitability better than the Boosted Tree model for
projects with only one team member, the complex blending methods would
take advantage of this interaction. In summary, in the complex methods,
the probability outputs from each model are interacted with each of the
original covariates. Model blending has the potential to elevate output
from the original methods by combining each of their strengths.

@@

The raw data from the case study company's database consisted of three
separate datasets: one describing the client and technical details for
each project, another with each employee's time sheet entries, and
another with all past invoices. This data was combined, cleaned, and
rearranged so that each project was described within one row. From the
timesheet data, 10 variables were engineered including the percent of
hours performed by `professional' employees as opposed to `technical'
employees, and the position of the employee that completed the most
hours on each project. 3 variables were engineered from the invoicing
data such as total amount invoiced for a project and mean invoice size
for a client. Finally, from the project dataset, 2 variables were
engineered including the number of projects completed with a client, and
project category. Text analysis of the project description determined
the project category variable. The response variable, whether the
project was profitable or not, was calculated by subtracting the cost of
employee hours from the timesheets from the sum of the paid invoices for
a project. The total amount invoiced for projects ranged from \$500 to
over \$1,000,000.

34 covariates made up the full dataset, and these were narrowed down to
11 in order to reduce predictive noise and provide a simple model that
is comprehendable for stakeholders (Weisberg 2005). Variable selection
methods included linear regression, random forests, and cforests and the
final variables are listed below.

\begin{itemize}
\tightlist
\item
  \% of Hours Completed by a Professional-level Employee
\item
  Time span
\item
  Number of Employees on Project
\item
  Amount Invoiced for Project
\item
  Business Category of Client
\item
  Project Category
\item
  Total Amount Invoiced from the Client - Past Jobs
\item
  Discipline
\item
  Position of Main Project Employee - new variable from ANOVA findings
\item
  \% of Hours by Main Project Employee
\item
  Billing Type
\end{itemize}

Timespan and the total amount invoiced are unknowns at before a project
commences. Therefore, these variables were discretised into 5 or 6
categories from which a manager would be able to select. The category
boundaries were determined by visualising clusters using hierarchical
dendrograms.

\subparagraph{3.2.1 Research Question 1}\label{research-question-1}

To answer the first research question, if loss making projects can be
predicted using statistical and machine learning algorithms, the binary
classification and blending methods were built and evaluated. The area
under the receiver operating characteristic (ROC) curve (AUC) statistic
was used to compare models and assess performance. Models were built on
a training subset of data (80\%) and validated on the remaining data.
Therefore, different training and testing sets produced different AUC
statistics. In order to compare which models performed significantly
better than others, an adequate sample size of AUC statistics was
required. A two-sample power calculation run on 20 AUC output for each
model determined that 100 samples were required to achieve a statistical
power of 0.8.

Then, the model output from the validation sets were combined and fed
into the three simple and three complex blending models. Again, 100
models were built to achieve a power of 0.8 for the variation in AUC
test statistics across different divisions of training/testing data. A
maximum of five training and validation divisions were created from a
complete set of individual model results.

\subparagraph{3.2.2 Research Question 2}\label{research-question-2}

The second research questions asks whether a predictive algorithm can
improve a company's profitability? This was analysed using a simple
decision-making scenario. The model with the highest AUC score may not
improve company profits the most, because this depends on \emph{which}
loss making and profitable projects are accurately being identified.
Therefore, all blended models as well as individual models were carried
forward for this analysis. A profit curve was built to compare methods,
which is a chart that plots the change in profit the company earns on
the y-axis vs.~the probability threshold on the x-axis. For this
decision making scenario, the threshold is the probability output value
(a value between 0 and 1 that the model predicts for each case), from
which it is decided whether a project will be accepted or rejected. For
example, if the output threshold is 0.6, and a model predicts a project
has probability of being a loss-making job of 0.7, the project is not
taken on by the case study company. All profits and losses from jobs
above the threshold were discounted. The y-axis of the profit curve is
the ratio of profit after rejecting projects based on model output to
historical profit:

\[
\begin{aligned}
Ratio\ of\ Model\ Profit\ to\ Historical\ Profit\ (\%) &= \frac{\sum_{p = 1}^{N} I(Pr_p < threshold) \cdot Profit_p}{\sum_{p = 1}^{N} Profit_p} \cdot100
\end{aligned}
\]

@@

If the threshold were zero, all jobs were rejected, the profit would be
\$0, and the ratio to historical profits would be 0\%. If the threshold
was 1.0, all jobs were accepted and the profit would be the same as
historical profits, giving a ratio of 100\%. The aim was to find the
optimal threshold point where mostly loss-making projects were rejected,
resulting in higher profits. 100 profit curves were built for each
method to achieve a power of 0.8 and a 95\% confidence interval could
also be determined around each threshold point. The final expected
increase in profit combined with the percentage of projects to be
rejected presents a clear scenario that the case study business managers
could strategically assess for implementation.

@@

@@

@@

@@

@@

\hypertarget{ref-Madigan1994}{}
Madigan, David, and Adrian E Raftery. 1994. ``Model Selection and
Accounting for Model Uncertainty in Graphical Models Using Occam's
Window.'' Journal Article. \emph{Journal of the American Statistical
Association} 89 (428): 1535--46.

@@

@@
